---
title: "EDA P1"
output: html_notebook
---

**Exploratory Data Analysis:**

```{r comics}
comics <-read.csv("comics.csv")

# Check levels of align
levels(comics$align)

# Check the levels of gender
levels(comics$gender)

# Create a 2-way contingency table
table(comics$align, comics$gender)

# Load dplyr
library(dplyr)

# Remove align level
comics <- comics %>%
  filter(align != "Reformed Criminals") %>%
  droplevels()
```

```{r comic_a}
# Load ggplot2
library(ggplot2)

# Create side-by-side barchart of gender by alignment
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "dodge")

# Create side-by-side barchart of alignment by gender
ggplot(comics, aes(x = gender, fill = align)) + 
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 90))

```

**Conditional Proportions:**

```{r comics_b}
tab <- table(comics$align, comics$gender)
options(scipen = 999, digits = 3) # Print fewer digits
prop.table(tab)     # Joint proportions
prop.table(tab, 2)  # Conditional on columns

# Plot of gender by align
ggplot(comics, aes(x = align, fill = gender)) +
  geom_bar()
  
# Plot proportion of gender, conditional on align
ggplot(comics, aes(x = align, fill = gender)) + 
  geom_bar(position = "fill")
```

**Distribution of One Variable:**

```{r comics_c}
# Change the order of the levels in align
comics$align <- factor(comics$align, 
                       levels = c("Bad", "Neutral", "Good"))

# Create plot of align
ggplot(comics, aes(x = align)) + 
  geom_bar()

# Plot of alignment broken down by gender
ggplot(comics, aes(x = align)) + 
  geom_bar() +
  facet_wrap(~ gender)

# Create barchart of flavor
ggplot(comics, aes(x = align)) + 
  geom_bar(fill = "chartreuse") + 
  theme(axis.text.x = element_text(angle = 90))
```

**Numerical Summaries:**

```{r cars}
cars <- read.csv("cars04.csv")

# Load package
library(ggplot2)

# Learn data structure
str(cars)

# Create faceted histogram
ggplot(cars, aes(x = city_mpg)) +
  geom_histogram() +
  facet_wrap(~ suv)

# No of Cylinders
carsunique(cars$ncyl)

# Filter cars with 4, 6, 8 cylinders
common_cyl <- filter(cars, cars$ncyl %in% c(4, 6, 8))

library(dplyr)

# Create hist of horsepwr
cars %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  ggtitle("All Cars")

# Create hist of horsepwr for affordable cars
cars %>% 
  filter(msrp < 25000) %>%
  ggplot(aes(x = horsepwr)) +
  geom_histogram() +
  xlim(c(90, 550)) +
  ggtitle(" Affordable Cars")
```

**Three binwidths:**
The binwidth determines how smooth your distribution will appear: the smaller the binwidth, the more jagged your distribution becomes. It's good practice to consider several binwidths in order to detect different types of structure in your data.

```{r cars_a}

# Create hist of horsepwr with binwidth of 3
cars %>%
  ggplot(aes(horsepwr)) +
  geom_histogram(binwidth = 3) +
  ggtitle("Binwidth of 3")

# Create hist of horsepwr with binwidth of 30
cars %>%
  ggplot(aes(horsepwr)) +
  geom_histogram(binwidth = 30) +
  ggtitle("Binwidth of 30")

# Create hist of horsepwr with binwidth of 60
cars %>%
  ggplot(aes(horsepwr)) +
  geom_histogram(binwidth = 60) +
  ggtitle("Binwidth of 60")

```

**Boxplots:**

In addition to indicating the center and spread of a distribution, a box plot provides a graphical means to detect outliers. We can apply this method to the msrp column (manufacturer's suggested retail price) to detect if there are unusually expensive or cheap cars.

```{r cars_b}
# Construct box plot of msrp
cars %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()

# Exclude outliers from data
cars_no_out <- cars %>%
  filter(msrp < 100000)

# Construct box plot of msrp using the reduced dataset
cars_no_out %>%
  ggplot(aes(x = 1, y = msrp)) +
  geom_boxplot()
```

**Plot Selection:**

Consider two other columns in the cars dataset: city_mpg and width. Which is the most appropriate plot for displaying the important features of their distributions? Remember, both density plots and box plots display the central tendency and spread of the data, but the box plot is more robust to outliers.

```{r cars_c}
# Create plot of city_mpg
cars %>%
  ggplot(aes(x = 1, y = city_mpg)) +
  geom_boxplot()

# Create plot of width
cars %>% 
  ggplot(aes(x = width)) +
  geom_density()

#Because the city_mpg variable has a much wider range with its outliers, it's best to #display its distribution as a box plot.
```

**MultiVariables:**

```{r cars_d}
# Facet hists using hwy mileage and ncyl
cars %>%
  ggplot(aes(x = hwy_mpg)) +
  geom_histogram() +
  facet_grid(ncyl ~ suv, labeller = label_both) +
  ggtitle("Mileage and Cycle")
```

```{r custom_unwanted}
# Filter for Asia, add column indicating outliers
gap_asia <- gap2007 %>%
  filter(continent == "Asia") %>%
  mutate(is_outlier = lifeExp < 50)

# Remove outliers, create box plot of lifeExp
gap_asia %>% 
  filter(!is_outlier) %>%
  ggplot(aes(x = 1, y = lifeExp)) +
  geom_boxplot()
```

