---
title: "Data Cleaning"
author: "Javeed Basha"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data and Library Loading

```{r data_library_load}
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
weather <- read.csv("weather.csv", stringsAsFactors = FALSE)
bmi <- read.csv("body_mass.csv", stringsAsFactors = FALSE)
bmi_cc <- read.csv("bmi_mass_cc.csv", stringsAsFactors = FALSE)
census <- read.csv("census_data.csv", stringsAsFactors = FALSE)
students <- read.csv("students_data.csv", stringsAsFactors = FALSE)
students2 <- read.csv("students_data_year.csv", stringsAsFactors = FALSE)
social_df <- read.csv("social.csv", stringsAsFactors = FALSE)
```


## Topics Covered:
1. Exploring Raw Data
2. Tidying Data
3. Analysing Data

## Investigating Raw Data

```{r dc_one}
# Data Views
head(weather)
tail(weather)
str(weather)

# Understanding the Data
class(bmi)
dim(bmi)
names(bmi) # Printing Cloumn Names
str(bmi)

glimpse(bmi)
summary(bmi)

# Printing
head(bmi, 10)
tail(bmi, 10)

# Plotting
hist(bmi$Y2008)
# Comparision Plot
plot(bmi$Y1980, bmi$Y2008)
```

## Tidying Data

Rows are referred as Observations.
Columns are referred as Variables.
One type of Observational Unit per table.

# Principles of Messy Data
1. Cloumn headers are values, not variable names.
2. Multiple variable are stored in a single column.
3. A single observational unit is stored in multiple tables.
4. Multiple types of observational unit are stored in the same table.

```{r data_two}
# Applying gather() - tidyr package to year columns
bmi_long <- gather(bmi, year, bmi_val, -Country)
# Checking dimensions
dim(bmi_long)

# Reversal spread()
bmi_wide <- spread(bmi, year, bmi_val)
head(bmi_wide)

# Separate Function
bmi_cc_clean <- separate(bmi_cc, col = Country_ISO, into = c("Country", "ISO"), sep = "/")
head(bmi_cc_clean)

# Unite Function
bmi_cc <- unite(bmi_cc_clean, Country_ISO, Country, ISO, sep = "-")
head(bmi_cc)

## Example - Column headers are values, not variable names
head(census, 10)
# Gather the month columns
census2 <- gather(census, month, amount, - YEAR)
# Arrange rows by YEAR using dplyr's arrange
census2 <- arrange(census2, YEAR)
# View first 20 rows of census2
head(census2, 20)

## Example - Variables are stored in both rows and columns
# Spread the type column
census_long2 <- spread(census_long, type, amount)
head(census_long2, 20)

## Example - Multiple values are stored in one column
census_long4 <- separate(census_long3, col = yr_month, 
                         into = c("year", "month"), sep = "_")
head(census_long4, 6)
```

## Analysing Data

```{r data_three}
## Analysing Students DataSet
str(students)
students$Grades <- as.character(students$Grades)
students$Medu <- as.factor(students$Medu)
students$Fedu <- as.factor(students$Fedu)
str(students)

# Lubridate Package Examples
str(students2)
# Parse as date
dmy("17 Sep 2015")
# Parse as date and time (with no seconds!)
mdy_hm("July 15, 2012 12:56")
# Coerce dob to a date (with no time)
students2$dob <- ymd(students2$dob)
# Coerce nurse_visit to a date and time
students2$nurse_visit <- ymd_hms(students2$nurse_visit)
str(students2)

## Working with Stringr Package
# Trim all leading and trailing whitespace
str_trim(c("   Filip ", "Nick  ", " Jonathan"))
# Pad these strings with leading zeros
str_pad(c("23485W", "8823453Q", "994Z"), width = 9, side = "left", pad = "0")

# Detect all dates of birth (dob) in 1997
str_detect(students2$dob, "1997")
# In the sex column, replace "F" with "Female"...
students2$sex <- str_replace(students2$sex, "F", "Female")
# ...And "M" with "Male"
students2$sex <- str_replace(students2$sex, "M", "Male")
# View the head of students2
head(students2)

## Missing Values
is.na(social_df)
any(is.na(social_df))
summary(social_df)

table(social_df$status)

# Example
social_df$status[social_df$status == ""] <- NA
complete.cases(social_df)
na.omit(social_df)

## Outliers & Errors
summary(students)

# View a histogram of the age variable
hist(students$age)
# View a histogram of the absences variable
hist(students$absences)
# View a histogram of absences, but force zeros to be bucketed to the right of zero
hist(students3$absences, right = FALSE)

# View a boxplot of age
boxplot(students3$age)
# View a boxplot of absences
boxplot(students3$absences)
```

## UseCase - Examples

```{r data_use_case}
# Exploration
class(weather)

dim(weather)

names(weather)

str(weather)

glimpse(weather)

summary(weather)

head(weather)

tail(weather)

# Tidying Data - Tidyr
weather2 <- gather(weather, day, value, X1:X31, na.rm = TRUE)
# Removing column of row names
weather2 <- weather2[, -1]

# Spread the data
weather3 <- spread(weather2, measure, value)
head(weather3)

## Preparation - Lubridate & Stringr
# Remove X's from day column
weather3$day <- str_replace(weather3$day, "X", "")
# Unite the year, month, and day columns
weather4 <- unite(weather3, date, year, month, day, sep = "-")
# Convert date column to proper date format using lubridates's ymd()
weather4$date <- ymd(weather4$date)
# Rearrange columns using dplyr's select()
weather5 <- select(weather4, date, Events, CloudCover:WindDirDegrees)
# View the head of weather5
head(weather5)


# Replace T with 0 (T = trace)
weather5$PrecipitationIn <- str_replace(weather5$PrecipitationIn, "T", 0)
# Convert characters to numerics
weather6 <- mutate_each(weather5, funs(as.numeric), CloudCover:WindDirDegrees)
str(weather6)

## Missing, Extreme, Outliers Data
# Count missing values
sum(is.na(weather6))
# Find missing values
summary(weather6)
# Find indices of NAs in Max.Gust.SpeedMPH
ind <- which(is.na(weather6$Max.Gust.SpeedMPH))
# Look at the full rows for records missing Max.Gust.SpeedMPH
weather6[ind, ]

# Review distributions for all variables
summary(weather6)
# Find row with Max.Humidity of 1000
ind <- which(weather6$Max.Humidity == 1000)
# Look at the data for that day
weather6[ind, ]
# Change 1000 to 100
weather6$Max.Humidity[ind] <- 100

# Look at summary of Mean.VisibilityMiles
summary(weather6$Mean.VisibilityMiles)
# Get index of row with -1 value
ind <- which(weather6$Mean.VisibilityMiles == -1)
# Look at full row
weather6[ind, ]
# Set Mean.VisibilityMiles to the appropriate value
weather6$Mean.VisibilityMiles[ind] <- 10

# Review summary of full data once more
summary(weather6)
# Look at histogram for MeanDew.PointF
hist(weather6$MeanDew.PointF)
# Look at histogram for Min.TemperatureF
hist(weather6$Min.TemperatureF)
# Compare to histogram for Mean.TemperatureF
hist(weather6$Mean.TemperatureF)
```

